\thispagestyle{empty}

\chapter*{Abstract}
Air pollution is one of the most pressing environmental challenges of the 21st century, with significant impacts on human health, ecosystems, and climate. Reliable forecasts are essential to support public policies, raise awareness, and inform daily decision-making. The Copernicus Atmosphere Monitoring Service (CAMS) provides high-quality forecasts of atmospheric composition worldwide, but existing platforms for accessing these data often involve a trade-off between scientific rigor and usability: expert-oriented tools require advanced technical skills, while public-facing applications simplify the information at the cost of analytical depth.

This work presents the design and implementation of online visualization platform that fills this void by providing a user-friendly interface to explore CAMS air quality forecasts. The system is entirely client-side and leverages open technologies such as Leaflet.js for mapping and Highcharts for dynamic time-series visualization. Key features include hourly forecast handling with an interactive time slider, pollutant value extraction at any clicked location, historical trend plotting, customizable color scales, and animation of forecast evolution. The platform automatically retrieves and processes CAMS WMS layers, ensuring scalability and adaptability to different pollutants and regions.

The results demonstrate that the developed platform successfully combines accessibility and analytical capability, enabling users to visualize and interpret complex atmospheric forecasts without specialized software or expertise. Beyond its academic value, this tool can support public awareness, educational purposes, and environmental monitoring initiatives. Future work will focus on integrating observational datasets for validation and incorporating advanced statistical analyses.


\clearpage
\thispagestyle{empty}
\hfill
\clearpage